---
output: pdf_document
graphics: yes
header-includes:
    - \usepackage{tabu}
    - \usepackage{amssymb, amsmath, amsthm}
    - \usepackage{enumerate}
    - \renewcommand{\P}{\textrm{P}}
    - \newcommand{\R}{\mathbb{R}}
    - \newcommand{\E}{\mathbb{E}}
    - \newcommand{\var}{{\rm Var}}
    - \newcommand{\cov}{{\rm Cov}}
    - \newcommand{\iid}{\stackrel{iid}{\sim}}
    - \newcommand{\N}{\mathcal{N}}
---
\noindent \begin{tabu} to \textwidth {@{}X[4 l] @{}X[r]}
  \textbf{Problem Set 3} 	       & \\ 
  \textbf{MFE 402: Econometrics} & \\ 
  \textbf{Professor Rossi}		   & 
\end{tabu}

This problem set is designed to review material on the multiple regression model and time series. Include both your R code and output in your answers.


## Question 1

Using a sequence of simple regressions computed in `R`, show how to obtain the multiple regression coefficient on $P2$ in the `multi` dataset from the `DataAnalytics` package. 

```{r}
library(DataAnalytics)
data("multi")
out = with(multi, lm(Sales~p1+p2))
out$coefficients[3]
```

## Question 2

Use matrix formulas and `R` code -- i.e., use `%*%` not `lm` -- to reproduce the least squares coefficients and standard errors shown on slide 17 of Chapter II. The `countryret` dataset is in the `DataAnalytics` package.

```{r}
data("countryret")
y = countryret$usa
X = cbind(rep(1, length(y)), countryret$canada, countryret$uk, countryret$australia,  countryret$france, countryret$germany,  countryret$japan)
b = round(chol2inv(chol(crossprod(X)))%*%crossprod(X,y), 5)

e = y - X%*%b
ssq = sum(e*e) / (length(y) - ncol(X))
Var_b = ssq*chol2inv(chol(crossprod(X)))
std_err = round(sqrt(diag(Var_b)), 5)

tab = data.frame(cbind(c('intercept','canada', 'uk', 'australia', 'france', 'germany', 'japan'), b, std_err))
colnames(tab) = c('','Estimate','Std Error')
tab
```

## Question 3

Run the regression of `VWNFX` on `vwretd`.

a. Compute a 90\% prediction interval for `VWNFX` when `vwretd` = 0.05 using the formulas in the class notes.
b. Check your work in part (a) by computing a 90\% prediction interval using R's `predict` command.

```{r}
library(reshape2)

data("Vanguard")
data("marketRf")
Van=Vanguard[,c(1,2,5)]
V_reshaped = dcast(Van, date~ticker, value.var = 'mret')
Van_market = merge(V_reshaped, marketRf, by='date')
X = Van_market$vwretd
y = Van_market$VWNFX

out = lm(y~X)

b_0 = out$coefficients[1]
b_1 = out$coefficients[2]
X_m = 0.05
Y_f = b_0 + b_1*X_m

N = length(out$residuals)
std = sqrt(sum(out$residuals^2/(N-2)))
s_spread = std*sqrt((1 + 1/N + (X_m - mean(X))^2/((N-1)*var(X))))
t_star = qt(0.05, N-2)
lwr = b_0 + b_1*X_m - abs(t_star)*s_spread
upr = b_0 + b_1*X_m + abs(t_star)*s_spread
cat(lwr, upr)
```

```{r}
predict(out, new = data.frame(X = 0.05), int = "prediction", level = 0.9)
```

## Question 4

Define the mean return vector and the symmetric variance-covariance matrix for 3 assets as follows:

$$ \mu = \begin{bmatrix} 0.010 \\ 0.015 \\ 0.025 \end{bmatrix} \hspace{3em} \Sigma = \begin{bmatrix} 0.0016 & 0.0010 & 0.0015 \\  & 0.0020 & 0.0019 \\  &  & 0.0042 \end{bmatrix} $$

a. Compute the correlation matrix of these three assets from the variance-covariance matrix $\Sigma$ by dividing the $(i,j)$ element of $\Sigma$ by $\sigma_i$ and $\sigma_j$. You must use matrix operations (e.g., `diag()`, `X*Y`, or `X%*%Y`) in your answer. You may not use a loop and you may not use the `R` function `cov2cor`. 
b. Compute the mean and standard deviation of a portfolio made from these assets with weights $(0.3, 0.4, 0.3)$

```{r}
Mu = matrix(c(0.010, 0.015, 0.025))
Sigma = matrix(c(0.0016, 0.0010, 0.0015, 0.0010, 0.0020, 0.0019, 0.0015, 0.0019, 0.0042), ncol=3)

sigma = diag(Sigma)

D_inverse = chol2inv(chol(sqrt(diag(sigma))))

Corr = D_inverse%*%Sigma%*%t(D_inverse)

# Corr = matrix(0, length(sigma), length(sigma))
# 
# for(row in 1:length(sigma)) {
#  for(col in 1:length(sigma)) {
#    Corr[row, col] = Sigma[row, col] / (sqrt(sigma[row]*sigma[col]))
#  }
# }

cat('a)\nCorrelation Matrix')
Corr


weights = matrix(c(0.3, 0.4, 0.3))
mean = t(weights) %*% Mu
std = sqrt(t(weights) %*% Sigma %*% weights)

cat('\nb)\nMean: ', mean, '\nStandard Deviation: ', std)
```

## Question 5

Using the same data as in Question 3 above and following the lecture slides (Chapter 3, section g), test the general linear hypothesis that $\beta_{up} = \beta_{down}$ in the following regression. Note that if you account for the `NA` values properly, you should get a slightly different result than what is presented in the lecture slides.

$$ VWNFX_t = \alpha + \beta_{up}*vwretd_t^{+} + \beta_{down}*vwretd_t^{-} + \varepsilon_t $$

```{r}
mkt_up=ifelse(Van_market$vwretd>0,1,0)
Van_market$upvw=mkt_up*Van_market$vwretd
Van_market$dwnvw=(1-mkt_up)*Van_market$vwretd
Van_market_sub = Van_market[, c(10,18,19)]
Van_market_sub = na.omit(Van_market_sub)
mkt_timing=lm(VWNFX~upvw+dwnvw,data=Van_market_sub)
lmSumm(mkt_timing)

R=matrix(c(0,1,-1),byrow=TRUE,nrow=1)
r=c(0)
X=cbind(c(rep(1,nrow(Van_market_sub))),Van_market_sub$upvw,Van_market_sub$dwnvw)
b=as.vector(mkt_timing$coef)
QFmat=chol2inv(chol(crossprod(X)))
QFmat=R%*%QFmat%*%t(R)
Violation=R%*%b-matrix(r,ncol=1)
fnum=t(Violation)%*%chol2inv(chol(QFmat))%*%Violation
n_minus_k = nrow(Van_market_sub)-length(b)
fdenom=nrow(R)*sum(mkt_timing$resid**2)/n_minus_k
f=fnum/fdenom
f

pvalue=1-pf(f,df1=nrow(R),df2=n_minus_k)
pvalue
```

## Question 6

Retrieve the Apple stock price series using the `quantmod` package (as done in the notes). Plot the autocorrelations of the difference in log prices.

```{r}
library(quantmod)
getSymbols('AAPL')
# head(AAPL)
lnP = log(AAPL$AAPL.Close)
acf(diff(lnP), na.action = na.omit)
```
